\begin{acknowledgments}

Many thanks to my supervisors, Harry Barrow \& David Young---not just
for their role in shaping this thesis, but for the education in general.

% -----------------------------------------------------------------------

I thank my friends who provided a good mix of help and distraction,
especially: Julian Budd, Teresa del Soldato, Inman Harvey, Jo Brook, Paulo
Costa, Ben du~Boulay, Geoff Goodhill, Suran Goonatilake, Richard Hall,
Sharon Groves, and Paul Booth.

Julia kept me sane while living through the best and worst of it
herself: from near-death in Pittsburgh to nights by the West Pier.
Thanks Jules.

A big thank-you to my family for all sorts of support on all sorts of
occasions.

% -----------------------------------------------------------------------

I specifically acknowledge: Mike McCloskey \& Grethe
Lindemann for their comments and hospitality;
Alistair
Chalmers for helping out
with some of the stats; Angela Jackson of the Education Development
Building for explaining how schools teach arithmetic; 
Kurt VanLehn for his comments; David Bennett for
explaining the BSB model; Judith Henstra at Experimental Psychology
for explaining a few things about eye movements;
and Shaaron
Ainsworth for
being, as she mistakenly called herself, ``the enemy''.

% -----------------------------------------------------------------------

This work was funded by a SERC award in conjunction with
Integral Solutions Ltd. Thanks to Julian Clinton and Alan Montgomery for
their support.  The simulations reported in this thesis were performed
with a modified
version of the \citeA{pdp3} {\em bp} program, and the analysis was
carried out in POPLOG POP-11.

\end{acknowledgments}
